### 2022-07-04 chapters/benchmarking/draft.tex
proofreading

\section{Introduction}
We are conducting empirical tests to determine the effect of programmer choice on software's energy consumption.
To accomplish this, one should not overlook the benchmarking pitfalls highlighted in the state of the art.
Moreover, when it comes to measuring the energy consumption of software, other obstacles arise, such as the impact of hardware, the difference between the machine used in the research and the machines that may apply the findings, etc. 
Moreover, significant improvements in computer science have resulted in an increase in the quantity of obsolete results. It is even more difficult to address all of them in studies that compare multiple solutions.
Furthermore, there may have been the appearance of new candidates as well as the development of others between the preliminary experiment and the published results. As a result, it is not only required to ensure the "reproducibility" of the results, but also to make way for new candidates. 
As a result, we'd like to propose a new concept for a fruitful experiment, which we'll call "Extension." The capacity to provide the tools required not only to repeat the experiment but also to add additional candidates, workloads, or crucial Key metrics.

@@

\subsection{Extension}

We propose making certain enhancements to the benchmarking framework that was suggested by the collaboratory on Experimental Evaluation of Software and Systems in Computer Science in order to address those problems \footnote{\url{http://evaluate.inf.usi.ch/}}.
Instead of only presenting the four primary aspects of their guidelines that were mentioned in state of the art\cref{sec:soa_reproducibility} we propose an abstract model to represent the experiment, 
\emph{context} : the hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers have a better judgement on the results and to be able reproduce the experiment while diminishing the impact of the external factor . 
An \emph{orchestrator} which its job to run the experiment. 
The orchestrator provides three intrafaces :

@@

\section{Reproducibility in benchmarking }
As we have discussed in the previuous section , the candidates shouldn't be related to the experiment, all they need to know is the interface of the workload, therefore. This section will discuss different ways to ensure a reproducibility of the experiment by comparing different ways to encapsulate the candidates

\subsection{Virtual machines}
First option should be using virtual machines,  Using virtual machines aka VM gives reaserchers the freedom to choose their own tools, software and operating system that they are the most confortable with without paying the price to change the actual working environment, which will give them eventually more controle over the dependencies and the execution environement. Moreover, using a vm will solve the \emph{replication crisis} thanks to the virtual images, even the most complex architecture can be reproduced easily by just instantiating a copy of the image. Since the virtual machines are agnostic to the host architecture, reaserchers won't have to worry about where and how their experiments are replicated because they have already setup the execution environement. Another advantage to the virtual machines is the snapshot mechanism, it allows reaserchers to create backups and revert some changes with simple clicks. Last but not least,thanks to the isolation, virtual machines push the reproducibility further by allowing the future usages to see all the variables -controled and uncontroled-  and do other analysis without dealing with any dependencies. In his paper \cite{howe_virtual_2012} bill howe lists the advantages of using virtual machines in reaserchers experements including the economical impact and cultural limitation to a such approach.
which allow them to have control over the ressouces, the dependencies and the execution environment. Moreover, thanks to the snapshots, deploying a software is easily done by instantiating a copy of that image.
However, this choice comes with a certain cost. Because the intervention of the hypervisor,the software will use two kernels , the virtual machine one and the host machine one, which will provide a noticiable overhead, and will impact the performances of the tests.Therefore, we can't use virtual machines for exepements that are related to performance. Another limiation with the virtual machine is the the isolation. It is true that this feature will prevent the experience environement with any undesirable interference from the outside world. but sometimes this contact is needed, especially when the experiment is dependent to an external part, such as sensors. In energicial tests we tend to use hardware powermeters which will make it difficult to use the virtual machines in this case .

% introduction of docker

\begin{figure}
    \center{\includegraphics[width=1\linewidth]{imgs/virtualization_techniques}}
    \caption{Different Methods of Virtualzation}\label{environement:virtualization_technique}
\end{figure}

\subsection{Containers}

another solution would be using something that allows us to have the isolation from the host os and the ease of replication that virtual machines offer, and the direct interaction with the hardware that the classical method give.
contarization offers a such advantages while keeping the isolation and the ease of replication for application.

Figure \ref{environement:virtualization_technique} explains the differents in architecture between the clasic types2 of Virtualisation and Containers.
\begin{itemize}
    \item Type 1: runs directly on the hardware, it is mainly used by the cloud providers where there is no main OS, but just virtaul machines, we can site for this the open-srouce XEN and VMware ESX
    \item Type 2: runs over the hostmachine Operating System, mostly used for personal comptuers, VMware server and virtualBox are famous examples of this type, most of the reaserchers expermentation are run witht this type, howerver due to the 2 Operatings syestms the applications tend to be more slower
    \item containers : Instead of its own kernel, containers used the hots kernel to run their Os, which makes them ligher, quickers and use the full pontial use of the hardware. For this we can cite \emph{Docker}, \emph{Linux LXC}\emph{LXD} \cite{abuabdo_virtualization_2019}
\end{itemize}
%  TODO Rephrase so i include it before VMs+ add links to the technologie

\subsection{Docker vs Virtual Machine}
Depsite that Type 1 is more performant than type 2, the second one is the most used in reaserch, since most researchers conducts their experements in their own machine. In the other hand, docker is the most famous thechonology for for containers.
In ower case we are more prone to docker for two reasons.
\begin{enumerate}
    \item we need a litetweight orchestrator to not affect the energy consumption of ower tests. As prior work mentioned [cite Morabito (2015) and van Kessel et al. (2016)]
          % cite Power efficiency of hypervisor-based virtuali- zation versus container-based virtualization. University of Amsterdam.
    \item since we are using the hardware itself to measure the energy consumption, we are required to interract with the host OS itself.
\end{enumerate}
Special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that allows us to retireve the energy consumption of a virtual machine.
%%%%% THINGS TO ADD : 
%% -------------- Dont think so 
% the impact of docker on the reprodudicibility 
% Limits of docker 
% The docker files and the clarification of the methodology 
\subsection{Docker and energy}
Now that we have chosen to go with the containers technorolgy to encapsulate our tests. What would be the impact of this solution on the energy consumption of our tests.

Based on the studies of \cite{santos2018does}, who analysed the impact of adding the docker layer on the energy consumption.
In their experement. Eddie Antonio et al run multiple benchmarks with and without docker. and compared their energy consumption and execution time.
The first step was to see the impact of docker deamon while there is no work. to see the impact of the orchestrator alone. Later they had the experiment with the following benchmarks
\begin{itemize}
    \item wordpress
    \item reddis
    \item postfresSQL
\end{itemize}
The following figures represents the energy consumption of the system while it is idle. As we can see  in figure \ref{fig:docker_idle} Docker braught arround 1000joules overhead.
\begin{figure}
    \center{\includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/idle_energy}}
    \caption{energy consumption of Idle system with and without docker \cite{santos2018does}}\label{fig:docker_idle}
\end{figure}
In the other hand as we can see in figure \ref{fig:docker_reddis}. docker increased the execution time of the benchmark by 50 seconds which caused an increase in energy since they are highly correlated.
The authors also highlated the fact that this increase of energy consumption is due to the docker deamon and not the fact that the application is in a container. Moreover they estimted the price of this extra energy and it was less than 0.15\$ in the worst case. Which is non significant compared to the advantages that docker bring for isolation and reproducibility.\\
if we  recap this study in one sentence,it would be the following one.
The dockerised softwares tend to consume more energy, because mainly they take more time to be executed.
The average power consumption is higher with only \textbf{2Watts} and it is due to the docker deamon.This overhead can be up to 5\% for IO intensive application, but it is mearly noticiable when it comes to CPU or DRMA intesive works

\begin{figure}
    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_time}
    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_energy}
    \caption{execution time and energy consumption of Redis  with and without docker \cite{santos2018does}}\label{fig:docker_reddis}
\end{figure}

\subsection{docker and accuracy}
And now Since the stat of the art has aggreed on the impact of docker on the energy consumption,Let's discuss it's impact on the accuracy. In other words\\
\textbf{RQ :} does Docker affect the energy variation of the exepements ?

To Answer this question we have conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.

>>>>

\section{Motivation}
Empirical measurements are critical tasks to capture the effect of developers' choice on software energy consumption.
To accomplish this, one should not overlook the benchmarking pitfalls highlighted in the state of the art.
Moreover, when it comes to measuring the energy consumption of software, other obstacles arise, such as the impact of hardware, the difference between the machines used for measurements purpose and the ones that may apply the findings, etc. 
Moreover, significant improvements in computer science have resulted in an increase in the quantity of obsolete results.
It is even more difficult to address all of them in studies that compare multiple solutions.
Furthermore, there may have been the appearance of new candidates, as well as the development of others between the preliminary experiment and the published results.
As a result, it is not only required to ensure the "reproducibility" of the results, but also to make way for new candidates.
Hence, we would like to propose a new concept to leverage experiments, which we call "Extension".
The capacity to provide the tools required not only to repeat the experiments, but also to include additional candidates, workloads, or relevant key performance indicators.

@@

\subsection{Extension}
We propose to extend the benchmarking framework suggested by the \emph{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}} to address those limitations.
Instead of only presenting the four primary aspects of their guidelines that were mentioned in state of the art~\cref{sec:soa_reproducibility}, we propose an abstract model to describe an empirical experiment:

\emph{context}: the hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers have a better judgement on the results and to be able reproduce the experiment while diminishing the impact of the external factor . 
An \emph{orchestrator} which its job to run the experiment. 
The orchestrator provides three interfaces:

@@

\section{Reproducibility in Benchmarking}
As discussed in the previous section, the candidates should not be related to the experiment, all they need to know is the interface of the workload.
This section thus discusses different ways to enforce the reproducibility of experiment by comparing different ways to encapsulate the systems-under-test.

\subsection{Virtual Machines}
First option should be using \emph{Virtual Machines} (VM), which gives researchers the freedom to choose the most appropriate tools, software and operating system that they are the most comfortable with, without paying the price to change the actual working environment, which will give them eventually more control over the dependencies and the execution environment.
Moreover, using a VM addresses the \emph{replication crisis} thanks to the virtual images, as even the most complex architecture can be reproduced easily by instantiating a copy of the image.
Since the virtual machines are agnostic to the host architecture, researchers will not have to worry about where and how their experiments are replicated, because they have already setup the execution environment.
Another advantage of VM is the snapshot mechanism, as it allows researchers to create backups and revert some changes with simple clicks. 
Last but not least, thanks to the isolation, virtual machines push the reproducibility further by allowing the future usages to see all the variables---controlled and uncontrolled---and do other analysis without dealing with any dependencies.
In his paper~\cite{howe_virtual_2012}, Bill Howe lists the benefits of using VM in researchers experiments, including the economical impact and cultural limitation to a such approach.
In particular, VM allow them to have control over the resources, the dependencies, and the execution environment.
Moreover, thanks to snapshots, deploying a software is easily achieved by instantiating a copy of that image.
However, this choice comes with a certain cost.
Because of the hypervisor, software will build on two kernels---the virtual machine (guest) one and the host machine one---which will provide a noticeable overhead, and will impact the performances of the system-under-test.
Therefore, we cannot use VM for experiments that are related to performance.
Another limitation of VM is the the isolation.
While this feature prevents the experimental environment from any undesirable interference from the outside world, this interaction may be required---especially when the experiment is dependent to an external part, such as sensors.
For energy measurements, one tend to use hardware power~meters, which will make it difficult to support VM in such a case.

% introduction of docker 
\subsection{Containers}
Another solution would be using something that allows us to benefit from the isolation of the host OS, while easing of replication proposed by VM, and the direct interaction with the hardware that the classical method give.
Containers offer such an advantage, while ensuring the isolation and the ease of replication for application.

Figure~\ref{environement:virtualization_technique} explains the differences in architecture between the types of virtualization and container technologies:
\begin{itemize}
    \item \textsf{Type\,1} runs directly on the hardware, it is mainly used by the cloud providers where there is no host OS, but just VM, use the open-source Xen or VMware ESX hypervisors to be executed;
    \item \textsf{Type\,2} runs over the host OS, mostly used for personal computers.
    VMware server and virtualBox are popular examples of this type, and most of the researchers experimentation are run with this type. However, due to the 2 OS, the applications tend to be slower.
    \item \textsf{Containers}. Instead of its own kernel, containers use the host kernel to run their OS, which makes them lighter, faster and use the full potential of hardware.
    For this category, one can cite \emph{Docker}, \emph{Linux LXC}, or \emph{LXD}~\cite{abuabdo_virtualization_2019}.
\end{itemize}
%  TODO Rephrase so i include it before VMs+ add links to the technologie

\begin{figure}
    \center{\includegraphics[width=1\linewidth]{imgs/virtualization_techniques}}
    \caption{Different Methods of Virtualization}\label{environement:virtualization_technique}
\end{figure}

\subsection{Docker Vs. Virtual Machine}
Despite the fact that \textsf{Type\,1} is more performant than \textsf{Type\,2}, the second one is the most used in research, as most researchers tend to conduct their experiments in their own machine.
In the other hand, Docker is the most famous container technology.
In our case, we are more prone to promote Docker for two reasons:
\begin{enumerate}
    \item we need a lightweight orchestrator to limit the overhead on energy consumption of our tests, as prior work mentioned [cite Morabito (2015) and van Kessel et al. (2016)],
          % cite Power efficiency of hypervisor-based virtuali- zation versus container-based virtualization. University of Amsterdam.
    \item as we are using hardware sensors to measure the energy consumption, we need to interact with the host OS.
\end{enumerate}

Special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that allows us to retireve the energy consumption of a virtual machine.
%%%%% THINGS TO ADD : 
%% -------------- Dont think so 
% the impact of docker on the reprodudicibility 
% Limits of docker 
% The docker files and the clarification of the methodology

\subsection{Docker \& Energy}
Now that we have chosen to go with the containers technology to encapsulate our tests.
What would be the impact of this solution on the energy consumption of our tests.

Based on the studies of \cite{santos2018does}, who analysed the impact of adding the Docker layer on the energy consumption.
In their experiment, Eddie Antonio~\emph{et~ al.} run multiple benchmarks with and without Ddocker. 
They compared the resulting energy consumption and execution time.
The first step was to see the impact of Docker deamon, while there is no work, to observe the impact of the orchestrator.
Then, they conduct their experiments with the following benchmarks:
\begin{itemize}
    \item Wordpress,
    \item Reddis,
    \item PostgreSQL.
\end{itemize}

The following figures represents the energy consumption of the system-under-test while it is idle.
As one can observe in Figure~\ref{fig:docker_idle}, Docker brings an overhead of around $1,000$ Joules.

\begin{figure}
    \center{\includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/idle_energy}}
    \caption{energy consumption of Idle system with and without docker \cite{santos2018does}}\label{fig:docker_idle}
\end{figure}

In the other hand, as one can see in Figure~\ref{fig:docker_reddis}.
Docker increased the execution time of the benchmark by 50 seconds, hence increasing the energy consumption.
The authors reported that this overhead is mostly due to the Docker deamon and not the fact that the application is in a container.
Moreover, they estimated the cost of this extra energy and it was less than 0.15\$ in the worst case, which is non-significant compared to the benefits that Docker brings for isolation and reproducibility.

To summarize, Docker-based software tends to consume more energy, because mainly they take more time to be executed.
The average power consumption is higher with only \textbf{2\,Watts}, due to the execution of the Docker deamon.
This overhead can reach up to 5\% for IO-intensive application, but it is merely noticeable when it comes to CPU- or DRMA-intesive workloads.

\begin{figure}
    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_time}
    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_energy}
    \caption{Execution time \& energy consumption of Redis with and without Docker~\cite{santos2018does}}\label{fig:docker_reddis}
\end{figure}

\subsection{Docker \& Accuracy}
As the state of the art assesses the impact of Docker on the energy consumption, one can also consider its impact on accuracy.
In other words:\\
\textbf{RQ:} does Docker affect the energy variation of the experiments? %La RQ n'est pas lié au reste du paragraphe, non ?

To answer this question we conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.

### 2022-07-04 chapters/introduction.tex
proofreading

\newpage
\chapter{Introduction}
\label{chapter:introduction}

Nowdays, computers are invasing our daily life from work to leasure, from fancy smartphone to an embded peacemaker that regulatres the heartbeat of some people. As human beigns we are known to use tools to enhance our bodies. And thanks to computers we pushed that step even further where now we are using machines to extend our brains. Some even argue that one day they will replace us, and therefore we created our own ending. Well this is the problem for the future generations. For the moment, the main concern of humanity is to keep this planet livebale until we find another alternative.

@@

Researchers are trying to solve this issue through different angles.
While some scientists are trying to find an alternative green source to generate energy. others focus on reducing it's%whose?
consumption. In computer science we are concerned with the ladder solution %reference.
Therefore most of the works are done on tunning the software and the hardware in ordrer to have a more efficient way use this energy. %more efficient way of producing programs from electricity? 
%say about hardware (i.e. superconductive materials with zero resistance)
In this thesis we are trying to find a way to make the energy consumption of the computer to be as low as possible. %on the software level
%More precisely, the purpose of this thesis is to highlight (attract attention to?) the impact of programers' choices on energy comsumption of software while running the code and to advise them with the strategies to write the greenest possible code.
Our approach is to reduce the energy consumption of the software services by changing certains parameters, such as palteform, programming langauge, and/or the design pattern.

The best way to do so is to formulate a theory behind the energy consumption of algorthms such as the complexity and the o notation.
Unfortunately this is not possible in the current state of the art. due to the lack of knowledge about the energy consumption of the algorithms, and the strong correlation between this consumption and the hardware configuration.

Unlike algrothim optimization in the field of perfornacne which is agnostic toward the plateform. the energy consumption of the algorithms is dependent on execution environement.
Therefore, for the moment we will start by formulating some hypthosis and explore them using empirical analysis.

Our work will be presented through the following chapters :

\begin{enumerate}
    \item \ref{chapter:literature_review}: Where we discuss the work done on the energy consumption and optimization in software engineering
    \item \ref{chapter:benchmarking}: It will present a set of guidelines and tools to help practitioners measure the energy consumption of their algorithms.
    \item : we will present the impact of programming languages on the energy consumption of the algorithms.
    \item : it will discuss the behaviour of python and the possibile ways to tune it in ordrer to reduce the energy consumpion
    \item : will present a study on java programming language and the impact of the JVM choice on the energy consumpion
    \item : as a prespective we introduce the impact of parallelisme on the enrgy consumption on time agnostic cases
\end{enumerate}

>>>>

\newpage
\chapter{Motivation}
\label{chapter:introduction}

Nowadays, computers are invading our daily life from work to leisure, from fancy smartphone to an embedded peacemaker that regulates the heartbeat of people.
As human beings, we are known to use tools to enhance our bodies. And thanks to computers we pushed that step even further where now we are using machines to extend our brains. Some even argue that one day they will replace us, and therefore we created our own ending. Well this is the problem for the future generations. For the moment, the main concern of humanity is to keep this planet liveable until we find another alternative.

@@

Researchers are trying to solve this issue through different angles.
While some scientists are trying to find an alternative green source to generate energy. others focus on reducing it's%whose?
consumption. In computer science we are concerned with the ladder solution %reference.
Therefore most of the works are done on tunning the software and the hardware in order to have a more efficient way use this energy. %more efficient way of producing programs from electricity? 
%say about hardware (i.e. superconductive materials with zero resistance)
In this thesis we are trying to find a way to make the energy consumption of the computer to be as low as possible. %on the software level
%More precisely, the purpose of this thesis is to highlight (attract attention to?) the impact of programers' choices on energy comsumption of software while running the code and to advise them with the strategies to write the greenest possible code.
Our approach is to reduce the energy consumption of the software services by changing certain parameters, such as platform, programming language, and/or the design pattern.

The best way to do so is to formulate a theory behind the energy consumption of algorithm, such as the complexity and the o notation.
Unfortunately this is not possible in the current state of the art. due to the lack of knowledge about the energy consumption of the algorithms, and the strong correlation between this consumption and the hardware configuration.

Unlike algorithm optimization in the field of performance, which is agnostic toward the platform, the energy consumption of the algorithms is dependent on execution environment.
Therefore, for the moment we will start by formulating some hypothesis and explore them using empirical analysis.

Our work will be presented through the following chapters:
\begin{enumerate}
    \item \ref{chapter:literature_review}: Where we discuss the work done on the energy consumption and optimization in software engineering
    \item \ref{chapter:benchmarking}: It will present a set of guidelines and tools to help practitioners measure the energy consumption of their algorithms.
    \item : we will present the impact of programming languages on the energy consumption of the algorithms.
    \item : it will discuss the behaviour of python and the possibile ways to tune it in ordrer to reduce the energy consumpion
    \item : will present a study on java programming language and the impact of the JVM choice on the energy consumpion
    \item : as a prespective we introduce the impact of parallelisme on the enrgy consumption on time agnostic cases
\end{enumerate}
